Redirecting Resolution Proofs

### 2011-05-24 bib/prooftransform/redirecting-proofs-2/paper.tex
minor fixes to text

\paragraph{Shorthand proofs.}
The last proof format we need to review is a shorthand notation for a subset of
Isar proofs. In their simplest form, these shorthand proofs are simply a list of
derivations, where `$a_1, \ldots, a_n \have c$' means ``from $a_1$ and $\ldots$
and $a_n$, we conclude $c$.'' If among the derivation's assumptions $a_j$
we have the previous derivation's conclusion, we can omit the assumption
and write `$a_1, \ldots, a_{j-1}, a_{j+1}, \ldots, a_n \hence c$' (corresponding
to Isar's \keyw{hence} and \keyw{thus} keywords). Depending on whether we use
the abbreviated format, our running example becomes
%
\begin{align*}
1, 2 & \have 4 &
    1, 2 & \have 4 \\
3, 4 & \have 5 &
    3 & \hence 5 \\
\B{0}, 4 & \have \B{6} &
    \B{0}, 4 & \have \B{6} \\
\B{0}, 5 & \have \B{7} &
    \B{0}, 5 & \have \B{7} \\
\B{6}, \B{7} & \have \bot &
    \B{6} & \hence \bot
\end{align*}
%
Each step is essentially a sequent $\Gamma \vdash \varphi$. The assumptions are
either the negated conjecture ($\B{0}$), facts that were proved elsewhere ($1$,
$2$, and $3$), or formulas that were proved in preceding sequents ($4$, $5$,
$\B{6}$, and $\B{7}$). The succedent of the last sequent is empty ($\bot$).

Direct proofs can be done the same way, but they have the constraints that we
may not assume the negated conjecture $\B{0}$ in any of the sequent and that the
last sequent has the conjecture $0$ as succedent. However, in some of the direct
proofs, we will find it useful to introduce case splits, as we had in the Isar
proof above. For example:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\frm 1, 2 \hav 4 \end
\frx 3 \hen 5 \end
\frm \hav 6, 7 \end
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&6] \\
      4 \hence {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&7] \\
      5 \hence {} & 0
    \end{aligned}$
  \end{tabular}}
\end{tabular}}\]
The notation in brace is a case split. In general, a case split has the form
\[\left[\begin{tabular}{l|c|l}
  $\begin{aligned}[t]
    [&\Gamma_1] \\
     \Delta_{11} \have {} & \varphi_{11} \\
     & ~\vdots \\
     \Delta_{1k_1} \have {} & \varphi_{1k_1}
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    ~\ldots~ \\
    ~\ldots~ \\
	\\
    ~\ldots~
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    [&\Gamma_n] \\
     \Delta_{n1} \have {} & \varphi_{n1} \\
     & ~\vdots \\
     \Delta_{nk_n} \have {} & \varphi_{nk_n}
    \end{aligned}$
\end{tabular}
\right]\]
with the requirement that a sequent with the succedent $\Gamma_1,
\ldots, \Gamma_n$ has been proved already. Also, each of the branches
must be a valid proof. The assumptions $[\Gamma_{\!j}]$ may be used to
discharge assumptions in the same branch, just as if they had been
sequents ${} \have \Gamma_{\!j}$. Seen from outside, the case split
expression stands for a sequent with the succedent
$\varphi_{1k_1}, \ldots, \varphi_{nk_n}$.

@@

We start with a simple proof by contradiction expressed as a proof tree and
using our shorthand notation:
%
\[\hbox{\begin{tabular}{l@{\qquad}l}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/linear-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0, 1 & \have \B 3 \\
2, \B 3 & \have \B 4 \\
1, \B 4 & \havebot
\end{aligned}$}
\end{tabular}}\]
Recall from Sect.\ \ref{sec:proof-notations} that the negated conjecture all all
the proof steps that are tainted by it are shown as negated. Next, we turn the
sequents around using contraposition at the sequent level to avoid all
negations. This gives
\begin{align*}
1, 3 & \have 0 \\
2, 4 & \have 3 \\
1 & \have 4
\end{align*}
Finally, if we reorder the sequents and introduce $\hencesym$ wherever possible,
we obtain a direct proof:

@@

The next two examples are lasso-shaped proofs:
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/lasso-1-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 0 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2 & \have \B 4 \\
\B 3, \B 4 & \have \B 5 \\
\B 5 & \have \B 6 \\
\B 6 & \havebot
\end{aligned}$}
&
\qquad\qquad
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/lasso-2-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 2 & \have \B 3 \\
\B 2 & \have \B 4 \\
\B 3 & \have \B 5 \\
\B 4 & \have \B 6 \\
\B 5, \B 6 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contradiction proof on the left-hand side. Starting from
$\bot$, it is easy to turn the proof around up to the lasso cycle:
%
\begin{align*}
& \have 6 \\
6 & \have 5 \\
5 & \have 3 \lor 4
\end{align*}
%
When applying the contrapositive to eliminate the negations in $\B 3, \B 4 \have
\B 5$, we obtain a disjunction on the right-hand side of the sequent: $5 \have 3
\lor 4$. To continue from there, we need a case split. Then we can finish then
branches:
%
\[\left[\begin{tabular}{l|l} %% TYPESETTING
    $\begin{aligned}[t]
      [&3] \\
      3 \have {} & 1 \\
      1 \have {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&4] \\
      4 \have {} & 2 \\
      2 \have {} & 0
    \end{aligned}$
  \end{tabular}\right]\]

The second lasso example is more tricky, because the cycle occurs near the end
of the contradiction proof. Already when redirecting the last inference, we get
a disjunction:
%
\begin{align*}
& \have 5 \lor 6
\end{align*}
%
And if we split naively and finish each branch independently of each other,
we end up with a fair share of duplication:
\pagebreak[3]
\[\left[\begin{tabular}{l|l} %% TYPESETTING
    $\begin{aligned}[t]
      [&5] \\
      5 \have {} & 3 \\
      3 \have {} & 2 \\
      2 \have {} & 1 \\
      1 \have {} & 0 \\
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&6] \\
      6 \have {} & 4 \\
      4 \have {} & 2 \\
      2 \have {} & 1 \\
      1 \have {} & 0 \\
    \end{aligned}$
  \end{tabular}\right]\]

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
 & [&5] & \\
{[2]} & $5 \have {}$ & $2 \lor 4$ & [6]
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more esthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.

@@

Once this procedure has terminated, we abbreviate case splits in which all
but one branch are trivial using ``rewriting,'' transforming
\[
\left[\begin{tabular}{c|c|c|r@{}l|c|c|c}
&&& $[$ & $c_i]$ &&& \\
&&& $\Delta_1 \have {}$ & $d_1$ &&& \\
&&&& $\vdots$ &&& \\
$[c_1]$ & $\cdots$ & $[c_{i-1}]$ & $\Delta_n \have {}$ & $d_n$ & $[c_{i+1}]$ & $\cdots$ & $[c_m]$
\end{tabular}
\right]
\]
into
\begin{align*}
\Delta_1 \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_1 \lor c_{i+1} \lor \cdots c_m \\
& \vdots \\
\Delta_n \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_n \lor c_{i+1} \lor \cdots c_m \\
\end{align*}
This works even if the $c_i$ branch has case splits. What we are doing
effectively is taking the nontrivial branch and adding $c_1, \ldots, c_{i-1},
c_{i+1}, \ldots, c_m$ as disjuncts in the conclusions.

>>>>

\paragraph{Shorthand proofs.}
The last proof format we need to review is a shorthand notation for a subset of
Isar proofs. In their simplest form, these shorthand proofs are simply a list of
derivations, where `$a_1, \ldots, a_n \have c$' means ``from $a_1$ and $\ldots$
and $a_n$, we conclude $c$.'' If among the derivation's assumptions $a_j$
we have the previous derivation's conclusion, we can omit the assumption
and write `$a_1, \ldots, a_{j-1}, a_{j+1}, \ldots, a_n \hence c$' (corresponding
to Isar's \keyw{hence} and \keyw{thus} keywords). Depending on whether we use
the abbreviated format, our running example becomes
%
\begin{align*}
1, 2 & \have 4 &
    1, 2 & \have 4 \\
3, 4 & \have 5 &
    3 & \hence 5 \\
\B{0}, 4 & \have \B{6} &
    \B{0}, 4 & \have \B{6} \\
\B{0}, 5 & \have \B{7} &
    \B{0}, 5 & \have \B{7} \\
\B{6}, \B{7} & \have \bot &
    \B{6} & \hence \bot
\end{align*}
%
Each step is essentially a sequent $\Gamma \vdash \varphi$. The assumptions are
either the negated conjecture ($\B{0}$), facts that were proved elsewhere ($1$,
$2$, and $3$), or formulas that were proved in preceding sequents ($4$, $5$,
$\B{6}$, and $\B{7}$). The succedent of the last sequent is $\bot$.

Direct proofs can be done the same way, but they have the constraints that we
may not assume the negated conjecture $\B{0}$ in any of the sequent and that the
last sequent has the conjecture $0$ as succedent. However, in some of the direct
proofs, we will find it useful to introduce case splits, as we had in the Isar
proof above. For example:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\frm 1, 2 \hav 4 \end
\frx 3 \hen 5 \end
\frm \hav 6, 7 \end
\cases{\begin{tabular}{l|l}
    $\begin{aligned}[t]
      [&6] \\
      4 \hence {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&7] \\
      5 \hence {} & 0
    \end{aligned}$
  \end{tabular}}
\end{tabular}}\]
The notation in brace is a case split. In general, a case split has the form
\[\left[\begin{tabular}{l|c|l}
  $\begin{aligned}[t]
    [&\psi_1] \\
     \Delta_{11} \have {} & \varphi_{11} \\
     & ~\vdots \\
     \Delta_{1k_1} \have {} & \varphi_{1k_1}
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    ~\ldots~ \\
    ~\ldots~ \\
	\\
    ~\ldots~
    \end{aligned}$\enskip\hbox{}&\enskip
  $\begin{aligned}[t]
    [&\psi_n] \\
     \Delta_{n1} \have {} & \varphi_{n1} \\
     & ~\vdots \\
     \Delta_{nk_n} \have {} & \varphi_{nk_n}
    \end{aligned}$
\end{tabular}
\right]\]
with the requirement that a sequent with the succedent $\psi_1 \lor
\ldots \lor \psi_n$ has been proved already. In addition, each of the branches
must be a valid proof. The assumptions $[\psi_{\!j}]$ may be used to
discharge assumptions in the same branch, just as if they had been
sequents ${} \have \psi_{\!j}$. Seen from outside, the case split
expression stands for a sequent with the succedent
$\varphi_{1k_1} \lor \ldots \lor \varphi_{nk_n}$.

@@

We start with a simple proof by contradiction expressed as a proof graph and
using our shorthand notation:
%
\[\hbox{\begin{tabular}{l@{\qquad}l}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/linear-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0, 1 & \have \B 3 \\
2, \B 3 & \have \B 4 \\
1, \B 4 & \havebot
\end{aligned}$}
\end{tabular}}\]
Recall from Sect.\ \ref{sec:proof-notations} that the negated conjecture all all
the proof steps that are tainted by it are shown as negated. Next, we turn the
sequents around using contraposition at the sequent level to avoid all
negations. This gives
\begin{align*}
1, 3 & \have 0 \\
2, 4 & \have 3 \\
1 & \have 4
\end{align*}
Finally, if we reorder the sequents and introduce $\hencesym$ wherever possible,
we obtain a direct proof:

@@

The next two examples are lasso-shaped proofs:
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/lasso-1-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 0 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2 & \have \B 4 \\
\B 3, \B 4 & \have \B 5 \\
\B 5 & \have \B 6 \\
\B 6 & \havebot
\end{aligned}$}
&
\qquad\qquad
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/lasso-2-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 2 & \have \B 3 \\
\B 2 & \have \B 4 \\
\B 3 & \have \B 5 \\
\B 4 & \have \B 6 \\
\B 5, \B 6 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contradiction proof on the left-hand side. Starting from
$\bot$, it is easy to turn the proof around up to the lasso cycle:
%
\begin{align*}
& \have 6 \\
6 & \have 5 \\
5 & \have 3 \lor 4
\end{align*}
%
When applying the contrapositive to eliminate the negations in $\B 3, \B 4 \have
\B 5$, we obtain a disjunction on the right-hand side of the sequent: $5 \have 3
\lor 4$. To continue from there, we need a case split. We can then finish the
branches:
%
\[\left[\begin{tabular}{l|l} %% TYPESETTING
    $\begin{aligned}[t]
      [&3] \\
      3 \have {} & 1 \\
      1 \have {} & 0
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&4] \\
      4 \have {} & 2 \\
      2 \have {} & 0
    \end{aligned}$
  \end{tabular}\right]\]

The second lasso example is more tricky, because the cycle occurs near the end
of the contradiction proof. Already when redirecting the last inference, we get
a disjunction:
%
\begin{align*}
& \have 5 \lor 6
\end{align*}
%
And if we split naively and finish each branch independently of each other,
we end up with a fair amount of duplication:
\pagebreak[3]
\[\left[\begin{tabular}{l|l} %% TYPESETTING
    $\begin{aligned}[t]
      [&5] \\
      5 \have {} & 3 \\
      3 \have {} & 2 \\
      2 \have {} & 1 \\
      1 \have {} & 0 \\
    \end{aligned}$\enskip\hbox{}&\enskip
    $\begin{aligned}[t] 
      [&6] \\
      6 \have {} & 4 \\
      4 \have {} & 2 \\
      2 \have {} & 1 \\
      1 \have {} & 0 \\
    \end{aligned}$
  \end{tabular}\right]\]

@@

Our final example is truly diabolical (and sightly unrealistic):
%
\[\hbox{\begin{tabular}{llll}
\raisebox{-\height}{\includegraphics[scale=0.45]{diagrams/diabolical-ref-graph.pdf}}
&
\qquad
\raisebox{-\height}{$\begin{aligned}[t]
\\
\B 0 & \have \B 1 \\
\B 1 & \have \B 2 \\
\B 1 & \have \B 3 \\
\B 2, \B 3 & \have \B 4 \\
\B 2, \B 4 & \have \B 5 \\
\B 3, \B 4 & \have \B 6 \\
\B 2, \B 5, \B 6 & \have \B 7 \\
\B 3, \B 6 & \have \B 8 \\
\B 7, \B 8 & \havebot
\end{aligned}$}
\end{tabular}}\]
%
We start with the contrapositive of the last sequent:
%
\[\have 7 \lor 8\]
%
Then we identify the nodes that dominate only one branch in the refutation graph
(i.e., that are reachable by navigating upward). Looking at the graph, we see
that $\B 5$ dominates $\B 7$ but not $\B 8$, and is the only such node. We
perform a case split and allow ourselves to perform inferences requiring 5 and 7
in the left branch and 8 in the right branch. First the right branch:
%
\begin{align*}
[&8] \\
8 \have {} & 3 \lor 6
\end{align*}
%
There is nothing more we can do for now: Any further inferences we would do in
the right branch would need to be repeated in the left branch. Now we do the
left branch:
%
\begin{align*}
[&7] \\
7 \have {} & 2 \lor 5 \lor 6
\end{align*}
We would now like to perform the inference $5 \have 2 \lor 4$. This requires
a case split:
%
\[\hbox{\begin{tabular}{@{}p{.5\textwidth}@{}p{.5\textwidth}@{}}
\cases{\begin{tabular}{l|r@{}l|l}
 & [&5] & \\
{[2]} & $5 \have {}$ & $2 \lor 4$ & [6]
\end{tabular}}\end{tabular}}\]
%
The 2 and 6 subbranches are left alone. Since only one branch of the subcase
split is nontrivial, we find it more aesthetically pleasing to abbreviate it to
%
\[2 \lor 5 \lor 6 \have 2 \lor 4 \lor 6\]
%
Hence, the left branch proves $2 \lor 4 \lor 6$, the right branch proves $3 \lor
6$, and both branches together prove $2 \lor 3 \lor 4 \lor 6$.

@@

Once this procedure has terminated, we abbreviate case splits in which all
but one branch are trivial using ``rewriting,'' transforming
\[
\left[\begin{tabular}{c|c|c|r@{}l|c|c|c}
&&& $[$ & $c_i]$ &&& \\
&&& $\Delta_1 \have {}$ & $d_1$ &&& \\
&&&& $\vdots$ &&& \\
$[c_1]$ & $\cdots$ & $[c_{i-1}]$ & $\Delta_n \have {}$ & $d_n$ & $[c_{i+1}]$ & $\cdots$ & $[c_m]$
\end{tabular}
\right]
\]
into
\begin{align*}
\Delta_1 \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_1 \lor c_{i+1} \lor \cdots c_m \\
& \vdots \\
\Delta_n \have {} & c_1 \lor \cdots \lor c_{i-1} \lor d_n \lor c_{i+1} \lor \cdots c_m \\
\end{align*}
This works even if the $c_i$ branch has nested case splits. We are effectively
taking the nontrivial branch and adding $c_1, \ldots, c_{i-1}, c_{i+1}, \ldots,
c_m$ as disjuncts in the conclusions.
